Downstream analyses of whole-genome bisulfite-sequencing data

### 2015-04-20 Rmarkdown/appendix.Rmd
Latex fixes
Complete revisions to WGBS downstream analysis.

\section{Chapter \ref{chap:wgbs_downstream_analyses}}

\subsection{Equivalence of tests to identify DMCs in a one-group experiment}\label{app:one_group_equivalence}

\subsubsection{Mathematical equivalence}

The saturated log-linear model can be written as $\log(F_{a, b}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{a, b}^{meth_state,sample}$, where $F_{a, b}$ is the value of the $(a, b)$ element in the $n \times 2$ contingency table. The null hypothesis that the cytosine is not a DMC is then tested by testing for an association between `sample` and `meth_state`, i.e., whether the parameter $\lambda_{a, b}^{meth_state, sample}$ is significantly different from zero.

This log-linear model can be reformulated as a logistic regression by noting that the log-odds (logits) for each row are $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \log(F_{a, 1}) - \log(F_{a, 2})$. Therefore,

\begin{align*}
\log(\frac{F_{a, 1}}{F_{a, 2}}) &= \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{1}^{sample} + \lambda_{a, 1} \\
&- (\lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{2}^{sample} + \lambda_{a, 2}) \\
&= \underbrace{(\lambda_{1}^{sample} - \lambda_{2}^{sample})}_{\alpha} + \underbrace{(\lambda_{a, 1} - \lambda_{a, 2})}_{\gamma_{i}}
\end{align*}

The expression $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \alpha + \gamma_{i}$ is a logistic regression model. We can test the null hypothesis that the cytosine is not a DMC by testing for a `sample`-effect on the log-odds, i.e., whether the parameter $\gamma_{i}$ is significantly different from zero.

Regardless of whether we use the log-linear model or the logistic model, to test the null hypothesis that the cytosine is not a DMC we can compute the deviance[^deviance] of the reduced model and compare it to a $\chi^{2}$ random variable with $(2 - 1) \times (n - 1)$ degrees of freedom. For the log-linear regression the reduced model is $\log(F_{a, b}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample}$ and for the logistric regression the reduced model is $\log(\frac{F_{a, 1}}{F_{a, 2}}) = \alpha$.

[^deviance]: The deviance is the log-likelihood ratio statistic comparing the current model within a saturated model.

\subsubsection{Computational equiavelence}

The computational equivalence of using Pearson's $\chi^{2}$ test, the deviance of a log-linear model and the deviance of a logistic regression are shown in the following R code:

```{r}
# The data
x <- cbind(M = c(38, 79, 59, 69, 44), U = c(1, 2, 1, 2, 46))
names(dimnames(x)) <- c("sample", "meth_state")
x

# Using Pearson's chi-square test
chisq.test(x)

# Fitting a log-linear model to perform iterative proportional fitting to the
# contingency table.
loglin(x, margin = list("sample", "meth_state"))
# An alternative formulation using loglm(), a convenient wrapper of loglin(),
# from the MASS package.
library(MASS)
loglm(formula =  ~ sample + meth_state, data = x)

# Fitting a generalised linear model with a binomial link using two different
# formulations
# (1) Using the array of frequencies
glm(x ~ 1, family = binomial)
# (2) Using relative frequencies (i.e., the beta-values) and weights (i.e.,
  # the sequencing coverage).
  glm(x[,1] / (x[, 1] + x[, 2]) ~ 1, family = binomial, weights = x[, 1] + x[, 2])
```
\section{Chapter \ref{chap:co-methylation_review}}\label{app:co-methylation_review}

@@

\subection{Estimate of number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}

>>>>

\chapter{Appendix}\label{chap:appendix}

\section{Chapter \ref{chap:co-methylation_review}}\label{app:co-methylation_review}

@@

\subsection{Estimate of number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}

### 2015-04-20 Rmarkdown/wgbs_downstream_analyses.Rmd
Complete revisions to WGBS downstream analysis.

A barrier to analyses based on m-tuples has been a lack of software. To overcome this barrier, I introduce \code{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples. \code{MethylationTuples} complements \code{methtuple} (Chapter \ref{chap:wgbs_bioinformatics_analysis}) by providing a framework for the manipulation and analysis of methylation patterns at m-tuples. I describe methods available in \code{MethylationTuples} for the downstream analysis of whole-genome bisulfite-sequencing data.
\end{chapabstract}

@@

We might estimate \widebar{B_{r}^{k}} by the group-wise average of the sample-wise weighted average of $\beta$-values for all methylation loci in the region, where the weights are proportional to the sequencing coverage. Then, identifying differential methylation at _a priori_ defined regions simply boils down to identifying differences in means, just as is the case for testing for DMCs. Again, this problem can be viewed as a 'stand-alone' test, such as a t-test, or framed as a regression problem to allow for the inclusion of additional covariates.

@@

I have also experienced considerable difficulty in applying some of these methods owing to poor software implementations. Two give two examples, `DMEAS` \citep{He:2013cj is only available as a Windows binary or as a Perl script that itself is only available as a PDF file and I have been unable to install `methclone` __TODO: Give reasons__.

@@

`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples, It is released under an Artistic-2.0 licence (__TODO: Why use this instead of MIT/GNU?__). I chose to write this software in R because it is a very popular language for data analysis, particularly in bioinformatics, and facilitates both batch and interactive usage. R is also my computational mother tongue and an R package is a convenient unit for sharing reusable code. To improve the performance of key functionality, parts of `MethylationTuples` are written in `C++`, making use of the wonderful `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.

@@

Bioconductor has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \citep{Lawrence:2013hi, Lawrence:2014gy}. However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges]. Therefore, I first wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 ([http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html](http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html)). In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations. For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., \code{type = `equal'}). Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.

@@

The `MethPat` object provides fast subsetting by rows (m-tuples) and columns (samples) via the `[` method. It also benefits from fast subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps()`-based methods. Several other useful utility functions are:

__UP TO HERE: Check these exist, update descriptions and add any new functions__

- `collapseStrand()`: Collapse strand-specific data by aggregating the counts. Only applicable to CpG methylation loci.
- `combine()`: Combine multiple `MethPat` objects into one.
- `filterOutVariants()`: Remove m-tuples that contain a known variant. Variants must be provided as a VCF.
- `getCoverage()`: Compute the sequencing coverage of the m-tuples in each sample.
- `IPD()`: Compute the IPD vector for each m-tuple.
- `methLevel()`: Compute $\beta$-values or $\mathcal{M}$-values. Only applicable to 1-tuples.
- `MBias()` and associated methods and functions: Helper functions to perform a more systematic analysis of M-bias (see Section \ref{sec:identifying_m-bias}).
- `plotBetaDensity()`: Plot the kernel density of the $\beta$-values for each sample, optionally stratified by a genomic feature, e.g., CpG islands.
- `tuples()`: Extract the $pos_{1}, \ldots, pos_{m}$ of the m-tuples.

@@

In the `MethPat` class, its associated methods and other utility functions, the `MethylationTuples` package provides a toolbox for manipulating methylation patterns at m-tuples. Aside from providing the necessary infrastructure to analyse methylation patterns at m-tuples, `MethylationTuples` currently includes specific methods to analyse and visualise co-methylation (Chapter \ref{chap:co-methylation}) with the `cometh` and `methLevelCor` functions. I also plan to add methods for estimating epialleles and epipolymorphism. And it is my hope that `MethylationTuples` will provide a useful foundation on which others can implement their own methods for analysing methylation patterns at m-tuples.

@@

The challenges of working with large datasets in R are well-known. These are in large part due to R being designed as an 'in-memory' application and its implementation of 'copy-on-modify' semantics \citep{wickham2014advanced}. More generally with large datasets, there is often a trade-off to be made between storage efficiency and algorithm simplicity --- a more efficient way of storing the data may be less convenient to work with and _vice versa_.

I have used the `MethylationTuples` package to analyses datasets containing up to 17 whole-genome bisulfite-sequencing samples (the Lister data), albeit using a server with $512$ Gb of (shared) memory. I have found the `MethPat` class to be a very convenient representation of the data, but one that does not coincide with the most efficient storage. This is more of a problem as the size of the m-tuples increases and the data become sparser. However, for values of $m < 5$, and particularly for 'dense' data, such as RRBS, this is far less of an issue. Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the EPISCOPE whole-genome bisulfite-sequencing data with various sized m-tuples. It also highlights the sparsity of the data as the size of the tuples grows --- most entries are $0$ (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample).

@@

The main inefficiency with the `MethPat` class is that the matrices storing the frequencies of each methylation pattern grow increasingly sparse as the size of the tuples increases. It may be possible to improve this by using a different approach to the internal storage of the data. For example, it may be possible to create an index so that the matrices storing the frequencies of methylation patterns can be re-ordered and stored as run-length encodings, a very space-efficient storage scheme.

In the short term, my aim for `MethylationTuples` is to release a version to Bioconductor with the core infrastructure for manipulating data of methylation patterns at m-tuples. In the longer term, I would like to extend the set of downstream analyses of m-tuples that are available in the package and for other users to do the same with their own methods.

>>>>

A barrier to analyses based on m-tuples has been a lack of software. To overcome this barrier, I introduce \texttt{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples. \texttt{MethylationTuples} complements \texttt{methtuple} (Chapter \ref{chap:wgbs_bioinformatics_analysis}) by providing a framework for the manipulation and analysis of methylation patterns at m-tuples. I describe methods available in \texttt{MethylationTuples} for the downstream analysis of whole-genome bisulfite-sequencing data.
\end{chapabstract}

@@

We might estimate $\widebar{B_{r}^{k}}$ by the group-wise average of the sample-wise weighted average of $\beta$-values for all methylation loci in the region, where the weights are proportional to the sequencing coverage. Then, identifying differential methylation at _a priori_ defined regions simply boils down to identifying differences in means, just as is the case for testing for DMCs. Again, this problem can be viewed as a 'stand-alone' test, such as a t-test, or framed as a regression problem to allow for the inclusion of additional covariates.

@@

I have also experienced considerable difficulty in applying some of these methods owing to poor software implementations. Two give two examples, `DMEAS` \citep{He:2013cj is only available as a Windows binary or as a Perl script that itself is only available as a PDF file and I have been unable to install `methclone` due to compilation errors.

@@

`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples. It is released under an Artistic-2.0 license, consistent with core Bioconductor packages. I chose to write this software in R because it is a very popular language for data analysis, particularly in bioinformatics, and facilitates both batch and interactive usage. R is also my computational mother tongue and an R package is a convenient unit for sharing reusable code. To improve the performance of key functionality, parts of `MethylationTuples` are written in `C++`, making use of the wonderful `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.

@@

Bioconductor has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \citep{Lawrence:2013hi, Lawrence:2014gy}. However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges]. Therefore, I first wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 ([http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html](http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html)). In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations. For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e., \texttt{type = `equal'}). Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.

@@

The `MethPat` object provides fast subsetting by rows (m-tuples) and columns (samples) via the `[` method. It also benefits from fast subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps()`-based methods. Several other useful utility functions for working with `MethPat` objects include:

- `collapseStrand()`: Collapse strand-specific data by aggregating the counts. Only applicable to CpG methylation loci.
- `combine()`: Combine multiple `MethPat` objects into one.
- `filterOutVariants()`: Remove m-tuples that contain a known variant. Variants must be provided as a `VCF` file.
- `findMTuples()`: Find m-tuples of a given size in a reference genome.
- `getCoverage()`: Compute the sequencing coverage of each m-tuples in each sample.
- `IPD()`: Compute the IPD vector for each m-tuple.
- `methLevel()`: Compute $\beta$-values or $\mathcal{M}$-values. Only applicable to 1-tuples.
- `tuples()`: Extract the $pos_{1}, \ldots, pos_{m}$ of the m-tuples.

@@

With the `MethPat` class, its associated methods and other utility functions, the `MethylationTuples` package provides a toolbox for manipulating methylation patterns at m-tuples. Aside from providing the necessary infrastructure to analyse methylation patterns at m-tuples, `MethylationTuples` currently includes specific methods to analyse and visualise co-methylation (Chapter \ref{chap:co-methylation}) with the `cometh()` and `methLevelCor()` methods. I also plan to add methods for estimating epialleles and epipolymorphism. It is my hope that `MethylationTuples` will provide a useful foundation on which others can implement their own methods for analysing methylation patterns at m-tuples.

@@

The challenges of working with large datasets in R are well-known. These are in large part due to R being designed as an 'in-memory' application and its implementation of 'copy-on-modify' semantics \citep{wickham2014advanced}. More generally, with large datasets, there is often a trade-off to be made between storage efficiency and algorithm simplicity; a more efficient way of storing the data may be less convenient to work with and _vice versa_.

I have used the `MethylationTuples` package to analyse various sized m-tuples from datasets containing up to 17 whole-genome bisulfite-sequencing samples (the _Lister_ data). I have found the `MethPat` class to be a very convenient representation of the data, however, it has also raised challenges that will apply for larger datasets.

One such challenge is the size of a `MethPat` object in RAM. The `MethPat` class currently favours a simpler implementation at the expense of storage efficiency. The main inefficiency with the `MethPat` class is that the matrices storing the frequencies of each methylation pattern grow increasingly sparse as the size of the tuples increases.

Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the EPISCOPE whole-genome bisulfite-sequencing data with various sized m-tuples. Wee see that as the size of the m-tuples increases, the data become sparser --- most counts are 0 (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample). However, for values of $m < 5$, and particularly for 'dense' data, such as RRBS, this is far less of an issue.

@@

It will be possible to improve the storage efficiency by using a different approach to the internal storage of the data. For example, it may be possible to use sparse matrices to store the counts or to create an index so that counts can be re-ordered and stored as run-length encodings, a very space-efficient storage scheme.

In the short term, my aim for `MethylationTuples` is to release a version to Bioconductor with the core infrastructure for manipulating data of methylation patterns at m-tuples. In the longer term, I would like to extend the set of downstream analyses of m-tuples that are available in the package and to help users extend the package to add their own methods.

\section{Summary}

Most methods for downstream analyses of bisulfite-sequencing data have focused on the problem of identifying differential methylation using methylation calls at 1-tuples. However, there is a growing interest in questions related to the heterogeneity of DNA methylation and these might be better addressed by analyses based on methylation patterns at m-tuples.

A barrier to these type of analyses has been a lack of software for extracting and manipulating these data. It is my hope that the `methtuple` and `MethylationTuples` software will prove useful in facilitating the development of methods for these new types of downstream analyses.
